---
title: "predicting lf"
author: "Owen Liu"
date: "November 13, 2015"
output: html_document
---

T
```{r packages,echo=F,results='hide'}
library(RColorBrewer)
library(ggplot2)
library(reshape2)
library(popbio)

#multiple plot function
multiplot <- function(..., plotlist=NULL, file, cols=1, layout=NULL) {
  library(grid)

  # Make a list from the ... arguments and plotlist
  plots <- c(list(...), plotlist)

  numPlots = length(plots)

  # If layout is NULL, then use 'cols' to determine layout
  if (is.null(layout)) {
    # Make the panel
    # ncol: Number of columns of plots
    # nrow: Number of rows needed, calculated from # of cols
    layout <- matrix(seq(1, cols * ceiling(numPlots/cols)),
                    ncol = cols, nrow = ceiling(numPlots/cols))
  }

 if (numPlots==1) {
    print(plots[[1]])

  } else {
    # Set up the page
    grid.newpage()
    pushViewport(viewport(layout = grid.layout(nrow(layout), ncol(layout))))

    # Make each plot, in the correct location
    for (i in 1:numPlots) {
      # Get the i,j matrix positions of the regions that contain this subplot
      matchidx <- as.data.frame(which(layout == i, arr.ind = TRUE))

      print(plots[[i]], vp = viewport(layout.pos.row = matchidx$row,
                                      layout.pos.col = matchidx$col))
    }
  }
}
```

## Introduction and Project Goals

This is an exploration of the use of matrix population models to explore
population dynamics in a metapopulation context, in which local populations
can be demographically open or closed, and vary in their demographic rates,
especially in the survival of YOY, or new recruits.  First, the model formulation
will be general, then I will attempt to parameterize it for specific species.

Much of the model formulation is based on Caswell (2001) and adapted from
Armsworth(2002). The initial goals (subject to change) are to:

* practice building matrix models

* investigate measurements of stability and resilience using such models

* think about the consequences of metapopulation structure for resilience

* practice using RMarkdown and GitHub

## Model Construction

### Parameters for the matrix model(s)

Age-structured matrix model, where population is structured into $\omega$ age classes, and abundance denoted by $\textbf{N} = (N_{1},...,N_{\omega})$. A vector $\textbf{p}$ describes survival, where, for example, $p_{i}$ is the proportional survival of individuals from age $i$ to age $i+1$.  Similarly, $\textbf{f}_{i}$ is the fertility of individuals of age $i$.  $\xi$ is the number of external larvae supplied to a local population; $a \in [0,1]$ is the self-recruitment rate, that is, the proportion of locally produced larvae that return to that population; and $\gamma$ is the proportion of settlers that remain alive at the end of the first year.

```{r params}
max.age <- 14 # max age class (number of age classes will be this plus 1 (for age 0)
gamma <- 0.125 # proportion of settling larvae that survive the first year
ext <- 1000 # number of externally supplied larvae
e.vec <- c(1,rep(0,(max.age))) # allows addition of external larvae to the matrix
ret <- 0.5 # proportion of locally produced larvae that return (self-seeding proportion)
surv <- 0.863 # natural survival of post-settler larvae
p.vec <- c(1,rep(surv,max.age)) # survivorship vector (p.vec[i] is the survival from age i-2 to i-1)

## Rate of self-recruitment (placeholder; sampled from a beta distribution later)
a <- 0.0016

## Von Bertalanffy parameters
Linf <- 52.2 #L infinity, asymptotic length
K <- 0.354 #growth rate
t0 <- -0.766 #theoretical age at zero length
vb <- c(Linf, K, t0) #concatenated 3 V-B growth params

## Fecundity/fertility/length relationship
repr.age <- 2 #age at first reproduction
f1 <- 0.0129
f2 <- 3.03

## Beverton-Holt parameter, for density-dependent model
B <- 1.1 # (B-H beta parameter)

## Years for each simulation
sim.years <- 200
```
#### Von Bertalanffy Growth

The classic Von Bertalanffy Growth equation, of the form

$$L(t) = L_\infty(1-e^{-K(t-t_0)})$$

```{r VB Growth}
vb.growth <- function(age.vec,vb)  { 
  ## function that takes an age vector and a vector of V-B params, c(Linf, K, t0)
  ## returns a vector of lengths
  lengths <- sapply(age.vec, function(x) vb[1]*(1-exp(-vb[2]*(x-vb[3]))))
  return(lengths)
  }

## with our parameters:
l.vec <- vb.growth(age.vec=0:max.age,vb=vb) #vector of lengths (to be used later)
plot(0:max.age,l.vec,main="Age vs. Length",type='l',xlab='Age, years',ylab='Length, cm',
     xaxs='i')
```

#### Fecundity relationship

This function relates length to fecundity, and is a power function that can be parameterized on a species-specific basis.  Adopted from Sadovy (1996)

*Note: fecundity is zero until reproductive age (a parameter above) is reached*

```{r fertility function}
 fert <- function(age.vec, vb, repr.age, f1,f2) {
  ## function that builds on VB equation above, taking three more parameters to calculate fertility
  ## as a function of age.
  ## From Armsworth (2002) and Sadovy (1996)
  l.vec <- vb.growth(age.vec=age.vec,vb=vb)
  f <- sapply(l.vec, function(x) f1*x^f2)
  f[1:repr.age] <- 0
  return(f)
}

# with our parameters:
f.vec <- fert(age.vec=0:max.age,vb=vb,repr.age=repr.age,f1=f1,f2=f2)
plot(0:max.age,f.vec,main="Age vs. Fecundity",type='l',xlab='Age, years',ylab='Fertility (# eggs)',
     xaxs='i')
```


#### Net Reproductive Rate

The net reproductive rate, $R$, is a useful concept, and is defined as the expected lifetime reproductive contribution from an individual that survives the first year, from Caswell (2001).  $R$ is useful because if $R\gamma$*a* (the expected per capita lifetime production of recruits) is less than 1, we can expect than the population will shrink (below replacement level; i.e. a *sink* population) because an individual is not expected to replace itself.  The converse is also true, that when $R\gamma$*a* is greater than 1, the population will be a *source* :

$$R = \sum_{t=1}^{\omega}f_{1}\prod_{k=0}^{t-1}p_{k}$$

```{r net repro rate}
net.R <- function(f.vec,p.vec) {
  p.cum <- numeric()
  for (i in 1:(max.age+1)) { #cumulative survival (i.e. vector of prob. surviving age 0:x)
    p.cum[i] <- prod(p.vec[1:i])
  }
  R <- sum(f.vec*p.cum)
  return(R)
}
```

### Local, density independent population

The density-independent model for a local population takes the form

$$\textbf{N}(\textit{t} + 1) = \textbf{MN}(\textit{t}) + \gamma\xi\textbf{e}$$

where **M** is a Leslie matrix, $\gamma$ is the proportion of settlers that survive the year, $\xi$ is the number of externally supplied larvae, and **e** is a vector to allow external larvae to effect the matrix. We will also use the defined parameter *a* $\in$ [0,1], which is the proportion of locally produced larvae that return to the same local adult population, i.e. the rate of self-recruitment.

We can now construct the Leslie matrix **M** :

```{r Leslie matrix density independent}
M.di <- function (max.age,a, gamma, f.vec, p.vec) {
  M=matrix(nrow=max.age+1,ncol=max.age+1)
  r1 <- f.vec*a*gamma #first row, allowing for a partially open population and settler mortality in year 1
  M[1,] <-r1
  for (i in 2:(max.age+1)) { #rows 2 to max age
    row <- numeric(max.age+1)
    row[i-1] <- p.vec[i]
    M[i,] <- row
  }
  return(M)
}

## a vector of stable age distribution (at equilibrium) can be calculated from the right eigenvector of the M matrix
## we will use this later to 'seed' simulations with appropriate abundances
stableage <- Re(eigen(M.di(max.age=max.age,a=a,gamma=gamma,f.vec=f.vec,p.vec=p.vec))$vector)[,1]
stableage <- stableage/sum(stableage)
stab.age.length <- data.frame(equil=stableage,age=0:14,len=l.vec)
ggplot(aes(x=age,y=equil),data=stab.age.length)+geom_point()
ggplot(aes(x=len,y=equil),data=stab.age.length)+geom_point()
```

Hypothetical length-frequency collection

```{r len freq}
#Make a stable age function (a and gamma assumed at their defaults f.vec for this exercise)
stab.age <- function(max.age=14,Linf=52.2,K=0.354,t0 = -0.766, surv=0.863, repr.age=2,f1=0.0129,f2=3.03) {
  vb <- c(Linf,K,t0)
  l.vec <- vb.growth(age.vec=0:max.age,vb=vb)
  f.vec <- fert(age.vec=0:max.age,vb=vb,repr.age=repr.age,f1=f1,f2=f2)
  p.vec <- c(1,rep(surv,max.age))
  agefreqs <- Re(eigen(M.di(max.age=max.age,a=0.0016,gamma=0.125,f.vec=f.vec,p.vec=p.vec))$vector)[,1]
  agefreqs <- agefreqs/sum(agefreqs)
  return(agefreqs)
}
bins <- seq(10,50,by=5)
dat <- dnorm(bins,mean=30,sd=8)
lf.rand <- data.frame(bin=bins,freq=dat)
ggplot(aes(x=len,y=equil),data=stab.age.length)+geom_point()+geom_point(aes(x=bin,y=freq*2.5),data=lf.rand,col='red',size=4)

Linf.sens <- list()
for(i in seq(35,60,by=5)) {
  data.temp <- stab.age(Linf=i)
  vb.temp <- c(i,K,t0)
  lvec.temp <- vb.growth(age.vec=0:max.age,vb=vb)
  Linf.sens[[paste(i)]] <- data.frame(Linf=i,len=lvec.temp,equil=data.temp)
}
Linf.sens<-melt(Linf.sens,id='Linf')
```
